---
title: Compound Annotation and Structure Collection Datasets
author: "Authors: Yuzhu Duan, Dan Evans, Jamison McCorrison, Nicholas Schork, Kevin Horan, Austin Leong, Siddharth Sai and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{compoundCollectionData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
fontsize: 14pt
bibliography: bibtex.bib
---

<!-- <style> -->
<!-- pre code { -->
<!--   white-space: pre !important; -->
<!--   overflow-x: scroll !important; -->
<!--   word-break: keep-all !important; -->
<!--   word-wrap: initial !important; -->
<!-- } -->
<!-- </style> -->

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('compoundCollectionData.Rmd', c('BiocStyle::html_document', 'pdf_document')); knitr::knit('compoundCollectionData.Rmd', tangle=TRUE)"
-->

# Overview

This package contains annotation and structure datasets for compounds in 
[DrugAge](https://genomics.senescence.info/drugs/), 
[DrugBank](https://www.drugbank.ca/), [CMAP02](https://portals.broadinstitute.org/cmap/) 
and [LINCS](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742) databases.
The description of the 5 datasets stored in this package is in the 
`Dataset Description` section. The compound annotation database is an SQLite 
database containing 5 tables, including 4 compound annotation tables for DrugAge, 
DrugBank, CMAP02 and LINCS databases, respectively. The other one is an ID 
mapping table of ChEMBL IDs to IDs of individual databases. The other 4 datasets 
stores the structures on drugs in the DrugAge, DrugBank, CMAP02 and LINCS 
databases in SDF files.

The actual datasets are stored in `AnnotationHub`. 

# Install and load package

## Installation and Loading
`compoundCollectionData` is a R/Bioconductor package and can be installed using 
`BiocManager::install()`.

```{r inst_bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("compoundCollectionData")
```

To obtain the most recent updates immediately, one can install it directly from 
GitHub as follows.
```{r inst_git, eval=FALSE}
devtools::install_github("yduan004/compoundCollectionData", build_vignettes=TRUE)
```

After the package is installed, it can be loaded into an R session as follows.
```{r load, eval=TRUE}
library(compoundCollectionData)
```

For detailed description of the package, please refer to the vignette by running
```{r, vig, eval=FALSE}
browseVignettes("compoundCollectionData")
```

# Dataset Description
The description of the 5 datasets in this package is as follows

## SQLite annotation database
It is a SQLite database storing compound annotation tables for DrugAge, DrugBank,
CMAP02 and LINCS, respectively. It also contains an ID mapping table of ChEMBL IDs
to IDs of individual databases. The following describes the columns and generation
of the 4 annotation tables.

The DrugAge database is manually curated by experts, it contains an extensive 
compilation of drugs, compounds and supplements (including natural products and 
nutraceuticals) with anti-aging properties that extend longevity in model organisms. 
In this package, the DrugAge database is downloaded from
[here](https://genomics.senescence.info/drugs/dataset.zip) as a CSV file. The 
downloaded `drugage.csv` file contains `compound_name`,	`synonyms`, `species`, `strain`,
`dosage`, `avg_lifespan_change`, `max_lifespan_change`,	`gender`, `significance`,
and `pubmed_id` annotation columns. Since the DrugAge database only contains the
drug name as identifiers, it is necessary to map the drug name to other uniform
drug identifiers, such as ChEMBL IDs. In this package,
the drug names have been mapped to [ChEMBL](https://www.ebi.ac.uk/chembl/),
[PubChem]((https://pubchem.ncbi.nlm.nih.gov/) and DrugBank IDs semi-manually
and stored under the `inst/extdata` directory named as `drugage_id_mapping.tsv`. 
Part of the id mappings in the `drugage_id_mapping.tsv` table is generated 
by the \code{processDrugage} function for compound names that have ChEMBL 
ids from the ChEMBL database (version 24). The missing IDs were added 
manually. A semi-manual approach was to use this 
[web service](https://cts.fiehnlab.ucdavis.edu/batch). After the semi-manual process,
the left ones were manually mapped to ChEMBL, PubChem and DrugBank ids. The 
entries that are mixture like green tee extract or peptide like Bacitracin were commented. 
This function builds the DrugAge annotation SQLite database from the 
Then the `drugage_id_mapping` table was built to the annotation SQLite database
named as `compoundCollection_0.1.db` by `buildDrugAgeDB` function.

The DrugBank annotation table was obtained from the downloaded DrugBank database
in [xml file](https://www.drugbank.ca/releases/latest).
The most recent release version at the time of writing this document is 5.1.5.  
The extracted xml file was processed by the \code{dbxml2df} function in this package.
`dbxml2df` and `df2SQLite` functions in this package was used to convert the xml 
file to data.frame and stored in the `compoundCollection` SQLite annotation database.
There are 55 annotation columns in the DrugBank annotation table, such as
`drugbank_id`, `name`, `description`, `cas-number`, `groups`, `indication`, 
`pharmacodynamics`, `mechanism-of-action`, `toxicity`, `metabolism`, `half-life`, 
`protein-binding`, `classification`, `synonyms`, `international-brands`, `packagers`, 
`manufacturers`, `prices`, `dosages`, `atc-codes`, `fda-label`, `pathways`, `targets`. 
The DrugBank id to ChEMBL id mappings were obtained from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src2.txt.gz).

The CMAP02 annotation table was processed from the downloaded compound 
[instance table](http://www.broadinstitute.org/cmap/cmap_instances_02.xls)
by the `buildCMAPdb` function in this package. The CMAP02 instance table contains
the following drug annotation columns: `instance_id`,	`batch_id`,	`cmap_name`,	`INN1`,	
`concentration (M)`, `duration (h)`, `cell2`,	`array3`, `perturbation_scan_id`, 
`vehicle_scan_id4`, `scanner`, `vehicle`, `vendor`, `catalog_number`, `catalog_name`. 
Drug names are used as drug identifies. The `buildCMAPdb` function maps the drug 
names to external drug ids including `UniProt`, `PubChem`, `DrugBank` and 
`ChemBank` ids. It also added additional
annotation columns such as `directionality`, `ATC codes` and `SMILES structure`.
The generated `cmap.db` SQLite database from `buildCMAPdb` function contains both
compound annotation table and structure information. The ChEMBL id mappings were
further added to the annotation table via PubChem CID to ChEMBL id mappings from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src22.txt.gz).
The CMAP02 annotation table was stored in the `compoundCollection` SQLite annotation
database. Then the CMAP internal IDs to ChEMBL id mappings were added to the ID 
mapping table. 

The LINCS compound annotation table was downloaded from 
[GEO](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE92nnn/GSE92742/suppl/GSE92742_Broad_LINCS_pert_info.txt.gz)
where only compounds were selected. The annotation columns are `lincs_id`, `pert_name`,
`pert_type`, `is_touchstone`, `inchi_key_prefix`, `inchi_key`, `canonical_smiles`, `pubchem_cid`.
The annotation table was stored in the `compoundCollection` SQLite annotation database. 
Since the annotation only contains LINCS id to PubChem CID mapping, the LINCS ids 
were also mapped to ChEMBL ids via inchi key.

The SQLite annotation database can be loaded into R as follows:
```{r sql, eval=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("compoundCollectionData", "annot_0.1"))
annot_path <- ah[["AH79563"]]
library(RSQLite)
conn <- dbConnect(SQLite(), annot_path)
dbListTables(conn)
dbDisconnect(conn)
```

## DrugAge SDF
It is an SDF (Structure-Data File) file storing molecular structures of DrugAge
compounds. The drug structures were obtained from PubChem CIDs from the DrugAge
annotation table by \code{getIds} function from \pkg{ChemmineR} package.

The DrugAge compound structures can be loaded into R as follows:
```{r da, eval=FALSE}
query(ah, c("compoundCollectionData", "drugage_build2"))
da_path <- ah[["AH79564"]]
da_sdfset <- ChemmineR::read.SDFset(da_path)
```

## DrugBank SDF
This SDF file stores structures of compounds in [DrugBank](https://www.drugbank.ca/)
database. The DrugBank SDF file was downloaded from 
[here](https://www.drugbank.ca/releases/latest#structures)
and made some validity checks and modifications via utilities in the `ChemmineR` package.

The DrugBank compound structures can be loaded into R as follows:
```{r db, eval=FALSE}
query(ah, c("compoundCollectionData", "drugbank_5.1.5"))
db_path <- ah[["AH79565"]]
db_sdfset <- ChemmineR::read.SDFset(db_path)
```

## CMAP SDF
The `cmap.db` SQLite database generated from the `buildCMAPdb` function
contains both compound annotation and structure information. Specifically, 
the `cmap.db` is loaded into R by the `initDb` function in the `ChemmineR` package,
then `getCompounds` function was used to get an `SDFset` object of 1,309 comopunds
in the CMAP02 database. The annotation table is stored in the datablock section 
of the SDFset. The SDFset was written to the CMAP02 SDF file by `write.SDF` function.

The CMAP02 compound structures can be loaded into R as follows:
```{r cmap, eval=FALSE}
query(ah, c("compoundCollectionData", "cmap02"))
cmap_path <- ah[["AH79566"]]
cmap_sdfset <- ChemmineR::read.SDFset(cmap_path)
```

## LINCS SDF
The LINCS compounds structures were obtained from PubChem CIDs in the annotation 
table via the `getIDs` function from the `ChemmineR` package.

The LINCS compound structures can be loaded into R as follows:
```{r lincs, eval=FALSE}
query(ah, c("compoundCollectionData", "lincs_pilot1"))
lincs_path <- ah[["AH79567"]]
lincs_sdfset <- ChemmineR::read.SDFset(lincs_path)
```

If further interested, The R script of generating the above 5 datasets is available 
at the `inst/scripts/make-data.R` file in this package.
