---
title: Integrating Community and Custom Compound Collections
author: "Authors: Yuzhu Duan, Dan Evans, Kevin Horan, Austin Leong, Siddharth Sai and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{customCMPdb}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
fontsize: 14pt
bibliography: bibtex.bib
editor_options: 
  chunk_output_type: console
---

<style>
pre code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('compoundCollectionData.Rmd', c('BiocStyle::html_document', 'pdf_document')); 
knitr::knit('compoundCollectionData.Rmd', tangle=TRUE)"
-->

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
  library(customCMPdb); library(ChemmineR)
})
```

# Overview

This package serves as a query interface for important community collections of
small molecules, while also allowing users to include custom compound
collections. Both annotation and structure information is provided.  The
annotation data is stored in an SQLite database, while the structure
information is stored in Structure Definition Files (SDF). Both are hosted 
on Bioconductor's `AnnotationHub`. A detailed description of the included 
data types is provided under the _Supplemental Material_ section of this vignette. 
At the time of writing, the following community databases are included: 

+ [DrugAge](https://genomics.senescence.info/drugs/) [@Barardo2017-xk]
+ [DrugBank](https://www.drugbank.ca/) [@Wishart2018-ap]
+ [CMAP02](https://portals.broadinstitute.org/cmap/) [@Lamb2006-du]
+ [LINCS](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742) [@Subramanian2017-fu]

In addition to providing access to the above compound collections, the package
supports the integration of custom collections of compounds, that will be
automatically stored for the user in the same data structure as the
preconfigured databases. Both custom collections and those provided by this
package can be queried in a uniform manner, and then further analyzed with
cheminformatics packages such as `ChemmineR`, where SDFs are imported into
flexible S4 containers [@Cao2008-np].

# Installation and Loading
As Bioconductor package `customCMPdb` can be installed with the 
`BiocManager::install()` function.
```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("customCMPdb")
```

To obtain the most recent updates of the package immediately, one can also install it 
directly from GitHub as follows.
```{r inst_git, eval=FALSE}
devtools::install_github("yduan004/customCMPdb", build_vignettes=TRUE)
```

Next the package needs to be loaded in a user's R session.
```{r load, eval=TRUE, message=FALSE}
library(customCMPdb)
library(help = "customCMPdb")  # Lists package info
```

Open vignette of this package.
```{r load_vignette, eval=FALSE, message=FALSE}
browseVignettes("customCMPdb")  # Opens vignette
```

# Overview
The following introduces how to load and query the different datasets.

## DrugAge Annotations
The compound annotation tables are stored in an SQLite database. This data can be 
loaded into a user's R session as follows (here for `drugAgeAnnot`).

```{r sql, eval=TRUE, message=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("compoundCollectionData", "annot_0.1"))
annot_path <- ah[["AH79563"]]
library(RSQLite)
conn <- dbConnect(SQLite(), annot_path)
dbListTables(conn)
drugAgeAnnot <- dbReadTable(conn, "drugAgeAnnot")
head(drugAgeAnnot)
dbDisconnect(conn)
```

## DrugAge SDF
The corresponding structures for the above DrugAge example can be loaded into an `SDFset` 
object as follows.
```{r da, eval=TRUE, message=FALSE, results=FALSE}
query(ah, c("compoundCollectionData", "drugage_build2"))
da_path <- ah[["AH79564"]]
da_sdfset <- ChemmineR::read.SDFset(da_path)
```

Instructions on how to work with `SDFset` objects are provided in the `ChemmineR` vignette 
[here](https://bioconductor.org/packages/ChemmineR/). For instance, one can plot any of 
the loaded structures with the `plot` function.

```{r da_chemmineR, eval=TRUE, message=FALSE, results=FALSE}
ChemmineR::cid(da_sdfset) <- ChemmineR::sdfid(da_sdfset[1:2])
ChemmineR::plot(da_sdfset[1])
```

## DrugBank SDF
The SDF from DrugBank can be loaded into R the same way.  The
corresponding SDF file was downloaded from
[here](https://www.drugbank.ca/releases/latest#structures). During the import
into R `ChemmineR` checks the validity of the imported compounds.

```{r db, eval=FALSE}
query(ah, c("compoundCollectionData", "drugbank_5.1.5"))
db_path <- ah[["AH79565"]]
db_sdfset <- ChemmineR::read.SDFset(db_path)
```

## CMAP SDF

The import of the SDF of the CMAP02 database works the same way.
```{r cmap, eval=TRUE, message=FALSE, results=FALSE}
query(ah, c("compoundCollectionData", "cmap02"))
cmap_path <- ah[["AH79566"]]
cmap_sdfset <- ChemmineR::read.SDFset(cmap_path)
```

## LINCS SDF

The same applies to the SDF of the small molecules included in the LINCS database.
```{r lincs, eval=TRUE, message=FALSE, results=FALSE}
query(ah, c("compoundCollectionData", "lincs_pilot1"))
lincs_path <- ah[["AH79567"]]
lincs_sdfset <- ChemmineR::read.SDFset(lincs_path)
```

For reproducibility, the R code for generating the above datasets is included
in the `inst/scripts/make-data.R` file of this package. The file location 
on a user's system can be obtained with `system.file("scripts/make-data.R", 
package="compoundCollectionData")`.

# Custom Annotation Database 
## Load Annotation Database
The SQLite Annotation Database is hosted on Bioconductor's `AnnotationHub`. 
Users can download it to a local `AnnotationHub` cache directory. The path to this
directory can be obtained as follows.
```{r download_db, eval=TRUE, message=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
annot_path <- ah[["AH79563"]]
```

## Add Custom Annotation Tables
The following introduces how users can import to the SQLite database 
their own compound annotation tables. In this case, the corresponding
ChEMBL IDs need to be included under the `chembl_id` column. 
The name of the custom data set can be specified under the `annot_name` 
argument. Note, this name is case insensitive. 
```{r custom, eval=TRUE, message=FALSE}
chembl_id <- c("CHEMBL1000309", "CHEMBL100014", "CHEMBL10",
               "CHEMBL100", "CHEMBL1000", NA)
annot_tb <- data.frame(cmp_name=paste0("name", 1:6),
        chembl_id=chembl_id,
        feature1=paste0("f", 1:6),
        feature2=rnorm(6))
addCustomAnnot(annot_tb, annot_name="myCustom")
```

## Delete Custom Annotation Tables
The following shows how to delete custom annotation tables
by referencing them by their name. To obtain a list of custom
annotation tables present in the database, the `listAnnot` function 
can be used.
```{r del, eval=TRUE, message=FALSE}
listAnnot()
deleteAnnot("myCustom")
listAnnot()
```

## Set to Default
The `defaultAnnot` function sets the annotation SQLite database back to the
original version provided by `customCMPdb`. This is achieved by deleting the
existing (e.g. custom) database and re-downloading a fresh instance from 
`AnnotationHub`.
```{r default, eval=FALSE}
defaultAnnot()
```

# Query Annotation Database
The `queryAnnotDB` function can be used to query the compound annotations from
the default resources as well as the custom resources stored in the SQLite
annotation database. The query can be a set of ChEMBL IDs. In this case it
returns a `data.frame` containing the annotations of the matching compounds
from the selected annotation resources specified under the \code{annot}
argument. The `listAnnot` function returns the names that can be assigned to
the `annot` argument.
```{r query, eval=TRUE, message=FALSE}
query_id <- c("CHEMBL1064", "CHEMBL10", "CHEMBL113", "CHEMBL1004", "CHEMBL31574")
listAnnot()
qres <- queryAnnotDB(query_id, annot=c("drugAgeAnnot", "lincsAnnot"))
qres
# query the added custom annotation
addCustomAnnot(annot_tb, annot_name="myCustom")
qres2 <- queryAnnotDB(query_id, annot=c("lincsAnnot", "myCustom"))
qres2
```

Since the supported compound databases use different identifiers, a ChEMBL
ID mapping table is used to connect identical entries across databases as
well as to link out to other resources such as ChEMBL itself or PubChem.  For
custom compounds, where ChEMBL IDs are not available yet, one can use
alternative and/or custom identifiers.
```{r not_chembl, eval=TRUE, message=FALSE}
query_id <- c("BRD-A00474148", "BRD-A00150179", "BRD-A00763758", "BRD-A00267231")
qres3 <- queryAnnotDB(chembl_id=query_id, annot=c("lincsAnnot"))
qres3
```

# Supplemental Material
## Description of Four Annotation Tables in SQLite Database
The DrugAge database is manually curated by experts, it contains an extensive 
compilation of drugs, compounds and supplements (including natural products and 
nutraceuticals) with anti-aging properties that extend longevity in model 
organisms [@Barardo2017-xk]. In this package, the DrugAge database is downloaded from
[here](https://genomics.senescence.info/drugs/dataset.zip) as a CSV file. The 
downloaded `drugage.csv` file contains `compound_name`, `synonyms`, `species`, `strain`,
`dosage`, `avg_lifespan_change`, `max_lifespan_change`, `gender`, `significance`,
and `pubmed_id` annotation columns. Since the DrugAge database only contains the
drug name as identifiers, it is necessary to map the drug name to other uniform
drug identifiers, such as ChEMBL IDs. In this package,
the drug names have been mapped to [ChEMBL](https://www.ebi.ac.uk/chembl/) [@Gaulton2012-ji],
[PubChem]((https://pubchem.ncbi.nlm.nih.gov/) [@Kim2019-tg] and DrugBank IDs semi-manually
and stored under the `inst/extdata` directory named as `drugage_id_mapping.tsv`. 
Part of the id mappings in the `drugage_id_mapping.tsv` table is generated 
by the \code{processDrugage} function for compound names that have ChEMBL 
ids from the ChEMBL database (version 24). The missing IDs were added 
manually. A semi-manual approach was to use this 
[web service](https://cts.fiehnlab.ucdavis.edu/batch). After the semi-manual process,
the left ones were manually mapped to ChEMBL, PubChem and DrugBank ids. The 
entries that are mixture like green tee extract or peptide like Bacitracin were commented.
Then the `drugage_id_mapping` table was built into the annotation SQLite database
named as `compoundCollection_0.1.db` by `buildDrugAgeDB` function.

The DrugBank annotation table was obtained from the downloaded DrugBank database
in [xml file](https://www.drugbank.ca/releases/latest).
The most recent release version at the time of writing this document is 5.1.5.  
The extracted xml file was processed by the \code{dbxml2df} function in this package.
`dbxml2df` and `df2SQLite` functions in this package were used to load the xml 
file into R and covert to a data.frame R object, then stored in the 
`compoundCollection` SQLite annotation database.
There are 55 annotation columns in the DrugBank annotation table, such as
`drugbank_id`, `name`, `description`, `cas-number`, `groups`, `indication`, 
`pharmacodynamics`, `mechanism-of-action`, `toxicity`, `metabolism`, `half-life`, 
`protein-binding`, `classification`, `synonyms`, `international-brands`, `packagers`, 
`manufacturers`, `prices`, `dosages`, `atc-codes`, `fda-label`, `pathways`, `targets`. 
The DrugBank id to ChEMBL id mappings were obtained from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src2.txt.gz).

The CMAP02 annotation table was processed from the downloaded compound 
[instance table](http://www.broadinstitute.org/cmap/cmap_instances_02.xls)
by the `buildCMAPdb` function in this package. The CMAP02 instance table contains
the following drug annotation columns: `instance_id`, `batch_id`, `cmap_name`, `INN1`,
`concentration (M)`, `duration (h)`, `cell2`, `array3`, `perturbation_scan_id`, 
`vehicle_scan_id4`, `scanner`, `vehicle`, `vendor`, `catalog_number`, `catalog_name`. 
Drug names are used as drug identifies. The `buildCMAPdb` function maps the drug 
names to external drug ids including `UniProt` [@The_UniProt_Consortium2017-bx], 
`PubChem`, `DrugBank` and `ChemBank` [@Seiler2008-dw] ids. It also adds additional
annotation columns such as `directionality`, `ATC codes` and `SMILES structure`.
The generated `cmap.db` SQLite database from `buildCMAPdb` function contains both
compound annotation table and structure information. The ChEMBL id mappings were
further added to the annotation table via PubChem CID to ChEMBL id mappings from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src22.txt.gz).
The CMAP02 annotation table was stored in the `compoundCollection` SQLite annotation
database. Then the CMAP internal IDs to ChEMBL id mappings were added to the ID 
mapping table. 

The LINCS compound annotation table was downloaded from 
[GEO](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE92nnn/GSE92742/suppl/GSE92742_Broad_LINCS_pert_info.txt.gz)
where only compounds were selected. The annotation columns are `lincs_id`, `pert_name`,
`pert_type`, `is_touchstone`, `inchi_key_prefix`, `inchi_key`, `canonical_smiles`, `pubchem_cid`.
The annotation table was stored in the `compoundCollection` SQLite annotation database. 
Since the annotation only contains LINCS id to PubChem CID mapping, the LINCS ids 
were also mapped to ChEMBL ids via inchi key.

# Session Info
```{r sessionInfo}
sessionInfo()
```

# References

