---
title: Customize and Query Compound Annotation Database
author: "Authors: Yuzhu Duan, Dan Evans, Jamison McCorrison, Nicholas Schork, Kevin Horan, Austin Leong, Siddharth Sai and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{customCMPdb}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
fontsize: 14pt
bibliography: bibtex.bib
editor_options: 
  chunk_output_type: console
---

<style>
pre code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('compoundCollectionData.Rmd', c('BiocStyle::html_document', 'pdf_document')); 
knitr::knit('compoundCollectionData.Rmd', tangle=TRUE)"
-->

```{r setup, echo=FALSE, messages=FALSE, warnings=FALSE}
suppressPackageStartupMessages({
  library(customCMPdb); library(ChemmineR)
})
```

# Overview

This package is served as the query and customization interface for compound 
annotations from [DrugAge](https://genomics.senescence.info/drugs/), 
[DrugBank](https://www.drugbank.ca/), [CMAP02](https://portals.broadinstitute.org/cmap/) 
and [LINCS](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE92742) databases.
It also stores the structure SDF datasets for compounds in the above four databases.

Specifically, the annotation database created by this package is an SQLite database
containing 5 tables, including 4 compound annotation tables from DrugAge [@Barardo2017-xk], 
DrugBank [@Wishart2018-ap], CMAP02 [@Lamb2006-du] and LINCS [@Subramanian2017-fu] databases, 
respectively. The other one is an ID mapping table of ChEMBL IDs to IDs of 
individual databases. The other 4 datasets stores the structures of compounds 
in the DrugAge, DrugBank, CMAP02 and LINCS databases in SDF files. The detailed 
description of the 5 datasets generated by this package is under the 
`Dataset Description` section of this vignette. The actual datasets are hosted 
in the `AnnotationHub`. 

This package also provides functionalities to customize and query the compound
annotation SQLite database. Users could add their customized compound annotation
tables to the SQLite database and query both the default (DrugAge, DrugBank, CMAP02,
LINCS) and customized annotations by providing ChEMBL ids of the query compounds.

# Installation and Loading
As Bioconductor package `customCMPdb` can be installed with the 
`BiocManager::install()` function.
```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("customCMPdb")
```

To obtain the most recent updates immediately, one can install it directly from 
GitHub as follows.
```{r inst_git, eval=FALSE}
devtools::install_github("yduan004/customCMPdb", build_vignettes=TRUE)
```

Next the package needs to be loaded into a user's R session.
```{r load, eval=TRUE, message=FALSE}
library(customCMPdb)
library(help = "customCMPdb")  # Lists package info
browseVignettes("customCMPdb")  # Opens vignette
```

# Dataset Description
The description of the 5 datasets in this package is as follows.

## SQLite annotation database
It is a SQLite database storing compound annotation tables for DrugAge, DrugBank,
CMAP02 and LINCS, respectively. It also contains an ID mapping table of ChEMBL IDs
to IDs of individual databases. The detailed description of the columns and generation
of the 4 annotation tables is available at the `Supplemental Material` section.  

The SQLite annotation database can be loaded into R as follows.
```{r sql, eval=TRUE, message=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("compoundCollectionData", "annot_0.1"))
annot_path <- ah[["AH79563"]]
library(RSQLite)
conn <- dbConnect(SQLite(), annot_path)
dbListTables(conn)
drugAgeAnnot <- dbReadTable(conn, "drugAgeAnnot")
head(drugAgeAnnot)
dbDisconnect(conn)
```

## DrugAge SDF
It is an SDF (Structure-Data File) file storing molecular structures of DrugAge
compounds. The drug structures were obtained from PubChem CIDs from the DrugAge
annotation table by \code{getIds} function from the \pkg{ChemmineR} package [@Cao2008-np]. 

The molecular structures of compounds in DrugAge database can be loaded into R as follows.
```{r da, eval=TRUE, message=FALSE, results=FALSE}
query(ah, c("compoundCollectionData", "drugage_build2"))
da_path <- ah[["AH79564"]]
da_sdfset <- ChemmineR::read.SDFset(da_path)
```

## DrugBank SDF
This SDF file stores structures of compounds in [DrugBank](https://www.drugbank.ca/)
database. The DrugBank SDF file was downloaded from 
[here](https://www.drugbank.ca/releases/latest#structures). Some validity checks 
and modifications were made via utilities in the `ChemmineR` package.

The molecular structures of compounds in DrugBank database can be loaded into R as follows.
```{r db, eval=FALSE}
query(ah, c("compoundCollectionData", "drugbank_5.1.5"))
db_path <- ah[["AH79565"]]
db_sdfset <- ChemmineR::read.SDFset(db_path)
```

## CMAP SDF
The `cmap.db` SQLite database generated from the `buildCMAPdb` function
contains both compound annotation and structure information. Specifically, 
the `cmap.db` is loaded into R by the `initDb` function in the `ChemmineR` package,
then `getCompounds` function was used to get an `SDFset` object of 1,309 compounds
in the CMAP02 database. The annotation table is stored in the datablock section 
of the SDFset. The SDFset was written to the CMAP02 SDF file by `write.SDF` function.

The molecular structures of compounds in CMAP02 database can be loaded into R as follows.
```{r cmap, eval=TRUE, message=FALSE, results=FALSE}
query(ah, c("compoundCollectionData", "cmap02"))
cmap_path <- ah[["AH79566"]]
cmap_sdfset <- ChemmineR::read.SDFset(cmap_path)
```

## LINCS SDF
The LINCS compounds structures were obtained from PubChem CIDs in the annotation 
table via the `getIDs` function from the `ChemmineR` package.

The molecular structures of compounds in LINCS database can be loaded into R as follows.
```{r lincs, eval=TRUE, message=FALSE, results=FALSE}
query(ah, c("compoundCollectionData", "lincs_pilot1"))
lincs_path <- ah[["AH79567"]]
lincs_sdfset <- ChemmineR::read.SDFset(lincs_path)
```

If further interested, The R script of generating the above 5 datasets is available 
at the `inst/scripts/make-data.R` file in this package. The file location can
be found by running `system.file("scripts/make-data.R",package="compoundCollectionData")` 
in user's R session or from the 
[GitHub repository](https://github.com/yduan004/customCMPdb/blob/master/inst/scripts/make-data.R) 
of this package.

# Customize Annotation Database 
## Load Annotation Database
The SQLite Annotation Database is hosted on `AnnotationHub`. Users could download
it to the local `AnnotationHub` cache foler and get its path as follows.
```{r download_db, eval=TRUE, message=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
annot_path <- ah[["AH79563"]]
```

## Add Custom Annotation Tables
The following shows functions used to add user's customized compound annotation 
tables to the annotation SQLite database. In this case, users need to know the 
corresponding ChEMBL ids of the added compounds and the added annotation table 
should contain a ChEMBL id column named as `chembl_id`. The `annot_name`
is a user defined name of the annotation table. It is case insensitive. 
If users want to add two different tables, the spells of the two `annot_name`
need to be different, not just the case of some characters.
```{r custom, eval=TRUE, message=FALSE}
chembl_id <- c("CHEMBL1000309", "CHEMBL100014", "CHEMBL10",
               "CHEMBL100", "CHEMBL1000", NA)
annot_tb <- data.frame(cmp_name=paste0("name", 1:6),
        chembl_id=chembl_id,
        feature1=paste0("f", 1:6),
        feature2=rnorm(6))
addCustomAnnot(annot_tb, annot_name="myCustom")
```

## Delete Custom Annotation Tables
The following shows the R code used to delete an added custom annotation resource
by providing the table name. `listAnnot` function could be used to list the 
available annotation resources in the SQLite annotation database.
```{r del, eval=TRUE, message=FALSE}
listAnnot()
deleteAnnot("myCustom")
listAnnot()
```

## Set to Default
The `defaultAnnot` function sets the annotation SQLite database to the default one
by deleting the existing database and re-downloading from `AnnotationHub`.
```{r default, eval=FALSE}
defaultAnnot()
```

# Query Annotation Database
The `queryAnnotDB` function can be used to query the compound annotations from 
the default resources as well as the custom resources stored in the SQLite annotation
database. The default annotation resources are `DrugAge`, `DrugBank`, `CMAP02`
and `LINCS`. Users customized compound annotations could be added/deleted as 
described above. 

The input of the query function could be a set of ChEMBL IDs, it returns a data.frame 
storing annotations of the input compounds from the selected annotation 
resources defined by the \code{annot} argument. The `listAnnot` function lists
the available options for the `annot` argument names.
```{r query, eval=TRUE, message=FALSE}
query_id <- c("CHEMBL1064", "CHEMBL10", "CHEMBL113", "CHEMBL1004", "CHEMBL31574")
listAnnot()
qres <- queryAnnotDB(query_id, annot=c("drugAgeAnnot", "lincsAnnot"))
qres
# query the added custom annotation
addCustomAnnot(annot_tb, annot_name="myCustom")
qres2 <- queryAnnotDB(query_id, annot=c("lincsAnnot", "myCustom"))
qres2
```

Since in the SQLite annotation database, ID identifiers from different ID systems, 
such as DrugBank and LINCS, are connected by ChEMBL IDs, it is hard to tell 
whether two IDs, such as DB00341, BRD-A42571354, refer to the same compound if 
either of them lack ID mappings to ChEMBL. So for querying compounds that don't 
have ChEMBL IDs, only one isolated database where the compounds belong to are 
supported. For example, a compound with LINCS id as "BRD-A00150179" doesn't have 
the ChEMBL ID mapping, when it is passed to the `chembl_id` argument of the 
`queryAnnotDB` function, the `annot` need only to be set as `lincsAnnot` and the result
will be the compound annotation table from the LINCS annotation.
```{r not_chembl, eval=TRUE, message=FALSE}
query_id <- c("BRD-A00474148", "BRD-A00150179", "BRD-A00763758", "BRD-A00267231")
qres3 <- queryAnnotDB(chembl_id=query_id, annot=c("lincsAnnot"))
qres3
```

# Supplemental Material
## Description of Four Annotation Tables in SQLite Database
The DrugAge database is manually curated by experts, it contains an extensive 
compilation of drugs, compounds and supplements (including natural products and 
nutraceuticals) with anti-aging properties that extend longevity in model 
organisms [@Barardo2017-xk]. In this package, the DrugAge database is downloaded from
[here](https://genomics.senescence.info/drugs/dataset.zip) as a CSV file. The 
downloaded `drugage.csv` file contains `compound_name`, `synonyms`, `species`, `strain`,
`dosage`, `avg_lifespan_change`, `max_lifespan_change`, `gender`, `significance`,
and `pubmed_id` annotation columns. Since the DrugAge database only contains the
drug name as identifiers, it is necessary to map the drug name to other uniform
drug identifiers, such as ChEMBL IDs. In this package,
the drug names have been mapped to [ChEMBL](https://www.ebi.ac.uk/chembl/) [@Gaulton2012-ji],
[PubChem]((https://pubchem.ncbi.nlm.nih.gov/) [@Kim2019-tg] and DrugBank IDs semi-manually
and stored under the `inst/extdata` directory named as `drugage_id_mapping.tsv`. 
Part of the id mappings in the `drugage_id_mapping.tsv` table is generated 
by the \code{processDrugage} function for compound names that have ChEMBL 
ids from the ChEMBL database (version 24). The missing IDs were added 
manually. A semi-manual approach was to use this 
[web service](https://cts.fiehnlab.ucdavis.edu/batch). After the semi-manual process,
the left ones were manually mapped to ChEMBL, PubChem and DrugBank ids. The 
entries that are mixture like green tee extract or peptide like Bacitracin were commented.
Then the `drugage_id_mapping` table was built into the annotation SQLite database
named as `compoundCollection_0.1.db` by `buildDrugAgeDB` function.

The DrugBank annotation table was obtained from the downloaded DrugBank database
in [xml file](https://www.drugbank.ca/releases/latest).
The most recent release version at the time of writing this document is 5.1.5.  
The extracted xml file was processed by the \code{dbxml2df} function in this package.
`dbxml2df` and `df2SQLite` functions in this package were used to load the xml 
file into R and covert to a data.frame R object, then stored in the 
`compoundCollection` SQLite annotation database.
There are 55 annotation columns in the DrugBank annotation table, such as
`drugbank_id`, `name`, `description`, `cas-number`, `groups`, `indication`, 
`pharmacodynamics`, `mechanism-of-action`, `toxicity`, `metabolism`, `half-life`, 
`protein-binding`, `classification`, `synonyms`, `international-brands`, `packagers`, 
`manufacturers`, `prices`, `dosages`, `atc-codes`, `fda-label`, `pathways`, `targets`. 
The DrugBank id to ChEMBL id mappings were obtained from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src2.txt.gz).

The CMAP02 annotation table was processed from the downloaded compound 
[instance table](http://www.broadinstitute.org/cmap/cmap_instances_02.xls)
by the `buildCMAPdb` function in this package. The CMAP02 instance table contains
the following drug annotation columns: `instance_id`, `batch_id`, `cmap_name`, `INN1`,
`concentration (M)`, `duration (h)`, `cell2`, `array3`, `perturbation_scan_id`, 
`vehicle_scan_id4`, `scanner`, `vehicle`, `vendor`, `catalog_number`, `catalog_name`. 
Drug names are used as drug identifies. The `buildCMAPdb` function maps the drug 
names to external drug ids including `UniProt` [@The_UniProt_Consortium2017-bx], 
`PubChem`, `DrugBank` and `ChemBank` [@Seiler2008-dw] ids. It also adds additional
annotation columns such as `directionality`, `ATC codes` and `SMILES structure`.
The generated `cmap.db` SQLite database from `buildCMAPdb` function contains both
compound annotation table and structure information. The ChEMBL id mappings were
further added to the annotation table via PubChem CID to ChEMBL id mappings from 
[UniChem](ftp://ftp.ebi.ac.uk/pub/databases/chembl/UniChem/data/wholeSourceMapping/src_id1/src1src22.txt.gz).
The CMAP02 annotation table was stored in the `compoundCollection` SQLite annotation
database. Then the CMAP internal IDs to ChEMBL id mappings were added to the ID 
mapping table. 

The LINCS compound annotation table was downloaded from 
[GEO](ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE92nnn/GSE92742/suppl/GSE92742_Broad_LINCS_pert_info.txt.gz)
where only compounds were selected. The annotation columns are `lincs_id`, `pert_name`,
`pert_type`, `is_touchstone`, `inchi_key_prefix`, `inchi_key`, `canonical_smiles`, `pubchem_cid`.
The annotation table was stored in the `compoundCollection` SQLite annotation database. 
Since the annotation only contains LINCS id to PubChem CID mapping, the LINCS ids 
were also mapped to ChEMBL ids via inchi key.

# Session Info
```{r sessionInfo}
sessionInfo()
```

# References

